A Mechanised Library for Generic (Co)Programming

### 2024-02-28 paper/paper.tex
Examples

In general not every functor has a fixed-point, so we are going to restrict
ourselves to polynomial functors as represented by containers.  A container is
defined by a type of \emph{shapes} and a \emph{family of position types} indexed
by shapes.
\begin{coqcode}
Context (Shape : Type) (Pos : Shape -> Type).
\end{coqcode}
In our previuos example, we would define \coq{Shape} as \coq{unit + A}
indicating there are two constructors in the data type and one of these contains
a piece of data of type \coq{A}. Moreover, we would define the positions
as \coq{Pos (inl tt) = Empty_set} and \coq{Pos (inr a) = unit} indicating that
the first constructor does not have any recursive type variables and the second
constructor has one type variable.

Therefore an \emph{extension} of this container is a functor defined as follows:
\begin{coqcode}
Record App (X : Type) := { shape : Shape; contents : Pos shape -> X }.
\end{coqcode}
with the obvious action on morphisms. Note that for \coq{Pos (inl tt)} we have a
type of functions \coq{Empty_set -> X} which is only inhabited by the trivial
map from the empty type and for \coq{Pos (inr a)} we have the type of
functions \coq{unit -> X} which has cardinality \coq{unit}.

Generally, any functor that is isomorphic
to a container extension is called a \emph{container functor}. However, in our
work we will call \emph{container functor} the functors defined by a specific
container extension.

The first problem in extracting Coq code from such container extensions is the
use of families of types. Since OCaml's type system cannot accurately capture
such families of types, Coq's code extraction mechanism will insert unsafe
casts. To avoid unsafe casts, we can consider representing the positions of a
container as a type that captures \emph{all} the potential positions in all
possible shapes, together with a predicate \coq{Valid} that
determines whether a position is valid in a shape.
\begin{coqcode}
Context (Shape : Type) (Pos : Type) (Valid : Shape -> Pos -> Prop).
Record App (X : Type) := { shape    : Shape; 
                          contents : {p | Valid shape p} -> X }.
\end{coqcode}
This approach solves the problem of the unsafe casts, since the code
extracted for \coq{{p | Valid shape pos}} will be an OCaml singleton
type defined as the OCaml equivalent to \coq{Pos}.

@@

\subsubsection{Polynomial Functors as Container Extensions}
Polynomial functors can be defined as container extensions~\ref{fig:poly}. We
omit here the proofs that these container extensions are isomorphic to
polynomials, but this proof is straightforward. To illustrate how the identity,
constant, sum and product container extensions work, consider the
tree-generating functor (i.e.\ the functor whose fixpoint is a binary tree):
$F\; X = \mathbb{N} + X \times X$.

@@

\section{Extracting Certified Hylomorphisms}

@@

@@

\bibliography{references.bib}
\appendix
\subsection{Course-of-Value Recursion with Finite Memory Table}
It can be proven that if $c$ is a recursive coalgebra, then the hylomorphism for
any algebra $a$ generates a finite output.  In other words, as far as the
hylomorphism is concerned $G_{\infty}$ can be regarded as its finite version
\[
  G_{*} A = \mu X. A \times G X
\]
We now rewrite course-of-value recursion using the finite memory table
For an algebra $a : D(D_{*} A) \to A$,
\begin{align*}
  & g : \mu F \to A
  & g = a \comp D(G_{*}(g) \comp \anacofree{\cata{G^{\lambda}(\InIso)}}) \comp \InOp
\end{align*}

>>>>

In general not every functor has a fixed-point, so we are going to restrict
ourselves to \emph{polynomial functors} as represented by containers.  A
container is defined by a type of \emph{shapes} and a \emph{family of position
types} indexed by shapes.
\begin{coqcode}
Context (Shape : Type) (Pos : Shape -> Type).
\end{coqcode}
For the functor \coq{F} we introduced in the previous paragraph we would
define \coq{Shape} as \coq{unit + A} indicating there are two constructors in
the data type and one of these contains a piece of data of type \coq{A}.
Moreover, we would define the positions as \coq{Pos (inl tt) = Empty_set}
and \coq{Pos (inr a) = unit} indicating that the first constructor does not have
any recursive type variables and the second constructor has one type variable.

At this point an \emph{extension} of this container is a functor defined as
follows:
\begin{coqcode}
Record App (X : Type) := { shape : Shape; contents : Pos shape -> X }.
\end{coqcode}
with the obvious action on morphisms given by post-composition
with \coq{contents}.  In our running example, the type \coq{App X}, for
some \coq{X} has two inhabitants.  The first is a pair composed
by \coq{inl tt : Shape} and a function \coq{Pos (inl tt) -> X}. This latter is
in fact a function of type we \coq{Empty_set -> X}  by definitions which in turn
is isomorphic to the type \coq{unit}, the type with only one inhabitant.  The
second inhabitant is the pair composed by \coq{inr a : Shape}, for
some \coq{a : A}, and a function \coq{Pos (inr a) -> X}.  This latter type is
equal to \coq{unit -> X} which represents the elements of \coq{X} and therefore
is isomorphic to \coq{X}.  This particular pair of shapes and positions is
isomorphic to the type \coq{unit + A * X}.

The correspondence between containers and polynomial functors has been
formalised in this work and has been moved to the
Appendix~\ref{sec:pol-containers} for space reasons. Generally, any functor that
is isomorphic to a container extension is called a \emph{container functor}.
Here we will call \emph{container functor} the functors defined
by a specific container extension.

\subsection{Extractable Containers}
In the previous section we defined containers using dependent types.  However,
Ocaml's type system is not equipped to handle these and to solve this problem
Coq's code extraction mechanism will insert unsafe casts.

Here consider instead representing the positions of a
container using a validity predicate which assigns shapes to positions.
A container extension is then defined as follows:
\begin{coqcode}
Context (Shape : Type) (Pos : Type) (Valid : Shape -> Pos -> Prop).
Record App (X : Type) := { shape    : Shape; 
                          contents : {p | Valid shape p} -> X }.
\end{coqcode}
This approach solves the problem of the unsafe casts, since the code extracted
for \coq{{p | Valid shape pos}} will be an OCaml singleton type defined as the
OCaml equivalent to \coq{Pos}.

@@

@@

\section{Certified Hylomorphisms}

@@

\subsection{Shortcut Deforestation}

\dcas{foldr/build examples from original paper (or hylo papers)}

@@

\bibliography{references.bib}
\appendix
\section{Polynomial Functors as Container Extensions}
\label{sec:pol-containers}
Polynomial functors can be defined as container extensions~\ref{fig:poly}. We
omit here the proofs that these container extensions are isomorphic to
polynomials, but this proof is straightforward. To illustrate how the identity,
constant, sum and product container extensions work, consider the
tree-generating functor (i.e.\ the functor whose fixpoint is a binary tree):
$F\; X = \mathbb{N} + X \times X$.

\section{Course-of-Value Recursion with Finite Memory Table}
It can be proven that if $c$ is a recursive coalgebra, then the hylomorphism for
any algebra $a$ generates a finite output.  In other words, as far as the
hylomorphism is concerned $G_{\infty}$ can be regarded as its finite version
\[
  G_{*} A = \mu X. A \times G X
\]
We now rewrite course-of-value recursion using the finite memory table
For an algebra $a : D(D_{*} A) \to A$,
\begin{align*}
  & g : \mu F \to A
  & g = a \comp D(G_{*}(g) \comp \anacofree{\cata{G^{\lambda}(\InIso)}}) \comp \InOp
\end{align*}
